The State of Urbit:\\Eight Years After the Whitepaper

### 2024-04-09 mss.tex
Include more refs.

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:1 (2024):  1–42. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Eight Years After the Whitepaper}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{The State of Urbit\\Eight Years After the Whitepaper}
\author{\authorname~\authorpatp \\ \affiliation}
\date{}

@@

\begin{abstract}
The Urbit whitepaper was released in 2016.  In the intervening eight years, the project has become much more fleshed out:  the sketch has started to take on color and detail.  This article explores the major developments since the whitepaper, including changes to the runtime, the network, the kernel, and the userspace.  It also discusses the current state of the project and intended refinements.
\end{abstract}

@@

The hard part of Kelvin shimming turned out to be that the current version of the userspace/kernelspace API is complex.  Taming that interface and localizing application-related state to make it easier to migrate will be critical for enabling Kelvin shims.  The shrubbery experiment is a step in this direction.

@@

The Urbit Improvement Proposal system was introduced in mid-2023.  This is an {\sc rfc} system based heavily on Bitcoin's {\sc bip} system and Ethereum's {\sc eip} system.  On a biweekly basis, proposals are reviewed by core developers in group calls.  Since its launch, over twenty UIPs have been written, of which thirteen have been ratified and built.

@@

On one hand, this representation allows for more expressivity than many comparable type systems at around the same spot on Barendregt's lambda cube (λω: types and values can both depend on types).  For example, the \lstinline[style=inlinecode]{$_} rune allows the programmer to declare the type of one expression in terms of the return type of any other expression, without that expression ever needing to be run, only ``type-inferred'' (synthesized).  This is very convenient for programs that return modified versions of themselves, which is ubiquitous in Urbit.

@@

The dotket rune (`.^`) no longer blocks until complete.  Because the agent model expects non-blocking IO, a single dotket could make an entire agent stuck if it can't be completed.  While a lot of work has gone into fleshing out the scry namespace, core devs are now entertaining the notion that the expectation of synchronous access to the namespace might not be worth it for agents.  An Urbit Improvement Proposal has been submitted that would remove the dotket rune in favor of a move-based interface to the namespace.

@@

As an optimization, the requesting ship places the first fragment of the command datum in the first request packet it sends.  In the common case of a 1KB command datum, the entire command+ack interaction consists of a single packet roundtrip.

@@

@@

All of the whitepaper's high-level claims about Hoon are still true, but a few changes to mold (type) declaration and the scope of the language's keyword equivalent (runes) has been expanded.\footnote{We consider the growth in Hoon to be somewhat unfortunate, and envision that it eventually slims back down somewhat as it is refined.}  There have been two or three revisions of syntax and type system.  The latest change added ``doccords'', a form of docstring, to the syntax.

@@

``Wills'' and ``deeds'', as described in this section, are no longer part of the system.  8-bit Galaxies, 16-bit stars, and 32-bit planets all have their public keys listed on Ethereum.  A 64-bit moon is assigned its public key by the planet that spawned and owns it, and other ships find the moon's public key by asking the planet.  A 128-bit comet signs its address with its own key, and other ships validate the comet's key by checking that it hashes to the comet's address.  This prevents comets from breaching, but it allows them to self-authenticate without being part of the on-chain {\sc pki}.

@@

Moving more network activity from writes to reads, through the scry namespace, has been shifting our conception of exactly-once delivery.  It now seems likely that by a year or two from now, Urbit's traditional notion of exactly-once delivery will be opt-in, built on top of the more general substrate of publishing scry bindings.

This requires another vaporware warning, but one could imagine a userspace application sending three poke commands to `~zod` by binding the commands to three paths like \lstinline[style=inlinecode]{/poke/~zod/foo/1}, \lstinline[style=inlinecode]{/poke/~zod/foo/2}, and \lstinline[style=inlinecode]{/poke/~zod/foo/3}, into a subtree of its own scry namespace.  The kernel would ensure \patp{zod} is notified of those bindings.  \patp{zod} would download the values at those paths and notify its \lstinline[style=inlinecode]{%foo} application.  The app would interpret the values as a sequence of poke commands, to be applied once and in order.  Once it processes each poke, it would bind an acknowledgment value to a predetermined path in its own namespace, such as \lstinline[style=inlinecode]{/ack/~nec/1}.

>>>>

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:1 (2024):  1–45. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Eight Years After the Whitepaper}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{The State of Urbit:\\Eight Years After the Whitepaper}
\author{\authorname~\authorpatp \\ \affiliation}
\date{}

@@

\begin{abstract}
\sloppy
The Urbit whitepaper was released in 2016.  In the intervening eight years, the project has become much more fleshed out:  the sketch has started to take on color and detail.  This article explores the major developments since the whitepaper, including changes to the runtime, the network, the kernel, and the userspace.  It also discusses the current state of the project and intended refinements.
\end{abstract}

@@

The hard part of Kelvin shimming turned out to be that the current version of the userspace/kernelspace {\sc api} is complex.  Taming that interface and localizing application-related state to make it easier to migrate will be critical for enabling Kelvin shims.  The shrubbery experiment is a step in this direction.

@@

The Urbit Improvement Proposal system was introduced in mid-2023.  This is an {\sc rfc} system based heavily on Bitcoin's {\sc bip} system and Ethereum's {\sc eip} system.  On a biweekly basis, proposals are reviewed by core developers in group calls.  Since its launch, over twenty {\sc uip}s have been written, of which thirteen have been ratified and built.

@@

On one hand, this representation allows for more expressivity than many comparable type systems at around the same spot on \citeauthor{Barendregt1991}'s lambda cube (λω: types and values can both depend on types).  For example, the \lstinline[style=inlinecode]{$_} rune allows the programmer to declare the type of one expression in terms of the return type of any other expression, without that expression ever needing to be run, only ``type-inferred'' (synthesized).  This is very convenient for programs that return modified versions of themselves, which is ubiquitous in Urbit.

@@

The \lstinline[style=inlinecode]{.^} dotket rune no longer blocks until complete.  Because the agent model expects non-blocking I/O, a single dotket could make an entire agent stuck if it can't be completed.  While a lot of work has gone into fleshing out the scry namespace, core devs are now entertaining the notion that the expectation of synchronous access to the namespace might not be worth it for agents.  An Urbit Improvement Proposal has been submitted that would remove the dotket rune in favor of a move-based interface to the namespace.

@@

As an optimization, the requesting ship places the first fragment of the command datum in the first request packet it sends.  In the common case of a 1~KB command datum, the entire command+ack interaction consists of a single packet roundtrip.

@@

Hoon keywords, employed in the whitepaper appendices, was removed due to disuse and disinterest.  Keyword Hoon may be re-introduced for pedagogical purposes, but for all essential purposes, runes are the most expressive and intuitive way to think about Hoon as a mature developer.

@@

All of the whitepaper's high-level claims about Hoon are still true, but a few changes to mold (type) declaration and the scope of the language's keyword equivalent (runes) has been expanded.\footnote{We consider the growth of Hoon to be somewhat unfortunate, and envision that it eventually slims back down somewhat as it is refined.}  There have been two or three revisions of syntax and type system.  The latest change added ``doccords'', a form of docstring, to the syntax.

@@

``Wills'' and ``deeds'', as described in this section, are no longer part of the system.  8-bit galaxies, 16-bit stars, and 32-bit planets all have their public keys listed on Ethereum.  A 64-bit moon is assigned its public key by the planet that spawned and owns it, and other ships find the moon's public key by asking the planet.  A 128-bit comet signs its address with its own key, and other ships validate the comet's key by checking that it hashes to the comet's address.  This prevents comets from breaching, but it allows them to self-authenticate without being part of the on-chain {\sc pki}.

@@

\sloppy
Moving more network activity from writes to reads, through the scry namespace, has been shifting our conception of exactly-once delivery.  It now seems likely that by a year or two from now, Urbit's traditional notion of exactly-once delivery will be opt-in, built on top of the more general substrate of publishing scry bindings.

\sloppy
This requires another vaporware warning, but one could imagine a userspace application sending three poke commands to `~zod` by binding the commands to three paths like \lstinline[style=inlinecode]{/poke/~zod/foo/1}, \lstinline[style=inlinecode]{/poke/~zod/foo/2}, and \lstinline[style=inlinecode]{/poke/~zod/foo/3}, into a subtree of its own scry namespace.  The kernel would ensure \patp{zod} is notified of those bindings.  \patp{zod} would download the values at those paths and notify its \lstinline[style=inlinecode]{%foo} application.  The app would interpret the values as a sequence of poke commands, to be applied once and in order.  Once it processes each poke, it would bind an acknowledgment value to a predetermined path in its own namespace, such as \lstinline[style=inlinecode]{/ack/~nec/1}.
